Union and intersection contracts: theory vs practice

### 2021-03-01 paper.tex
Skeleton and first few parts
WIP

\section{First section}
Some text

\section{Second section}
Some text

\section{Third section}
Some text

\section*{Conclusion}
Conclusion

>>>>

\section{Background on union types}

\subsection{Nickel design space}

\color{red}Goals of the section

Introduce Nickel, explain the choice of gradual
typing and main design orientations, such as practice-oriented, lightweight,
etc.\vspace{0.5cm}\color{black}

This paper is centered around the design of the Nickel language. Nickel is a
configuration programming language, meaning that a Nickel program eventually
evaluates to a value that can be directly serialized to JSON, YAML or alike.
Nickel aims at being lightweight, easy to write and easy to embed in a larger
system, to serve either directly as a native language to specify configuration,
or as a meta-language to generate configuration. The focus is first and foremost
practical: Nickel originated as a rethink of the language of the Nix package
manager, used to describe the biggest set of software packages ever built [ ],
and targets build systems and cloud deployment tools (Kubernetes, Terraform or
NixOps) as well.  While there's no hesitation in incorporating recent research
when it serves a well-defined purpose, the design decisions are guided by the
industrial use cases.

Our aim is that writing simple Nickel should not be more complex than writing
JSON or YAML. Unlike JSON, though, Nickel anticipates large configurations by
being both programmable and typed. In the configuration setting, there is a
singular dilemma with respect to typing: if your program is only run once on
fixed inputs, as are such programmable configurations, any type error will be
reported at run-time anyway. Why bother with the complexity of a static type
system? On the other hand, more and more software systems offload complexity to
configurations, such as cloud infrastructure, build system, or operating
systems, leading the corresponding configurations to become increasingly
complex, to the point where static types are beneficial. For reusable code —
that is, library functions — static types brings all the usual, widely
acknowledged benefits.

This dilemma is naturally solved by gradual typing [], an approach capable of
reuniting static typing and dynamic typing. Unlike gradually-typed language like
TypeScript though, we do not seek to build a complex type system that tries hard
to type most of the idioms naturally arising in dynamically typed code.  Rather,
we choose to provide a reasonably expressive type system with good inference
properties, that should make writing typed generic function that operate on data
relatively easy, but may require resorting to untyped code for more exotic expressions.
However, this is not a surrender, as in fact Nickel provides an alternative and
complementary mechanism for more complex data validation, contracts[references].

\subsection{Unions types}
\color{red}Goals of the section

Introduce union in general, and most usual form in statically typed languages, tagged unions\vspace{0.5cm}\color{black}

In programming, unions are fundamental and ubiquitous building blocks of data,
and Nickel is no exception. While products represent conjunctions, which are several pieces of
data stored all together in the same value, unions represent disjunctions, which are
several exclusive alternatives stored in one value. When nested together, unions
and intersections are able to describe a large class of data shapes: lists,
trees, structured errors, and so on [ref, examples?].

\begin{verbatim}
EXAMPLES of union in configuration specification
\end{verbatim}

In statically typed languages, and more specifically in functional programming
languages, unions are usually embodied by algebraic data types [references],
also called \textbf{tagged unions}. Tagged means that there is an additional
field, or constructor, that allows to discriminate between each alternative.
When an union is constructed, the programmer specifies which alternative is
chosen by using the corresponding tag. When a union is consumed, usually
via pattern matching, the code branches depending on the tag, and each branch is
ensured to handle data of the right shape. Well-known programming languages with
first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
Swift and surely many more.

\begin{verbatim}
EXAMPLES
\end{verbatim}

\subsection{Tagged versus untagged}
\color{red}Goals of the section

Interpreted languages already need to tag their values, hence bare union are more natural.
It also corresponds better to prior usage in untyped code\vspace{0.5cm}\color{black}

In the statically typed setting, the actual representation of a tagged union is
an implementation detail: from the point of view of the programmer, a tagged
union is a fist-class data type associated with constructors and destructors all
with well-defined semantics.  However, in a gradually typed language like
Nickel, the question of the representation of tagged unions as untyped values
arises.

Unfortunately, it seems that tagged unions do not have a canonical counterpart
as serializable untyped data. A natural representation is a record with a tag
and a value which shape depends on the tag:

\begin{lstlisting}
{ tag: <A, B, C>, value: Dyn }
\end{lstlisting}

A better way to think of it is maybe as dependent $\Sigma$-type which first
component is a tag and second component is the data type corresponding to this
tag:
\[
    \Sigma_{tag : <A, B, C>} \text{Alternative}(\text{tag})
\]

Alas, choosing this representation as a default is not fully satisfying. As any
representation, it does not map precisely to a JSON type. In consequence, a lot
of illegal combinations are structurally valid as untyped JSON values. More
importantly, this does not necessarily match with the configuration schemas one
can find in the wild, that favor simple, untagged union [example of JSON
schemas/protobuf/whatever with unions].

In dynamically typed languages, which are interpreted for the vast majority
[ref/examples?], it is customary to directly store different kind of data in one
variable without further ceremony. The runtime system needs to attach type tags
to values anyway, which can usually be observed using functions like the
JavaScript's \verb+typeof+. In practice, using so-called \textbf{untagged
unions} has already been an idiom in dynamically typed language for a long time
[ref/examples], and being able to type and handle these unions is an important
aspect of the migration to a gradual type system. Indeed, most of gradually
typed languages either existed originally as dynamically typed language or aim
at backward compatibility with a dynamically typed base language. This is why
they tend to favor supporting untagged union types, the typed counterpart of
this practice [example of gradually typed language with untagged union].

In the case of Nickel, there is no pre-existing untyped language that we have to
support, so we are freer of our choice. But are we? As a configuration language,
Nickel does not evolve in the abstract. It aims at the best interoperability
possible with JSON. It is designed to generate data for other softwares to
consume, and tries as much as possible to be expressive enough to encode data
schemas at the language level. If these schemas use untagged union, then this is
a strong incentive to support them at the type level, or at least, at the
contract level.

\subsection{Union types}

\subsection{Tagged as untagged}
\color{red}Goals of the section

Untagged unions, modulo a slightly smarter typechecker, can encode usual type-safe
tagged unions. They let the user free of the representation though, and encode
more stuff, that's why they are appealing.\vspace{0.5cm}\color{black}

Untagged union, together with the other existing features of the type system,
are sufficient to implement the encoding mentioned in [cite previous section]:

Encoding as: use singleton contract for tags, and then union.
\begin{lstlisting}
type sum = A of Num | B of Str
\end{lstlisting}

\begin{lstlisting}
{ tag : <A>, val : Num} | { tag : <B>, val : Str}
\end{lstlisting}

There is a twist, though. As it is, the static type system is unable to refine
the type of the value, making the following example rejected:

\begin{lstlisting}
switch x.tag {
  A => x.val + 1,
  B => then strings.fromNum x.val
}
\end{lstlisting}

Flow typing[ ], as implemented for example by TypeScript, is capable of solving
this issue by correctly propagating the type information in each branch. One can
imagine doing the same in Nickel. Thus equipped with untagged unions and a
typechecker smart enough, classic tagged unions can be simulated, but with the
benefit that the language is not imposing any representation. Untagged unions
empower the programmer to represent and handle a wider variety of unions
appearing in the wild. Thus, adding union types (and their natural and
unavoidable counterpart, intersections types [ref]) to Nickel sounds appealing.

\subsection{Union and intersections}
\color{red}Goals of the section

Explain why having unions probably unavoidably leads one to have intersection.\vspace{0.5cm}\color{black}

\section{Contracts}
\color{red}Goals of the section

A primer on contracts, which in the end are the problematic bit when confronted
with unions.\vspace{0.5cm}\color{black}
